Towards multiversal modal operators for homotopy type theory

### 2023-11-30 slides-hottest2023.lagda.tex
Slides finished

\newcommand{\dnote}[1]{%
  \begin{tabular}{@{}m{2em}@{}m{0.80\textwidth}@{}}%
    \textdbend &#1%
  \end{tabular}%
  \par
}

@@

  \textbf{Examples.} (classically)\ $(\NN,{\leq}),\ \ X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X).$
  \bigskip
  \pause

@@

\begin{frame}{Sequences provided by \textsc{lem}}
  \justifying
  \begin{block}{}
    \justifying
    \textbf{Lemma.} Let~$X$ be a well quasiorder. Let~$\alpha : \NN \to X$ be a
    sequence. Assuming~\bad{\textsc{lem}}, there merely is a monotonic
    subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
  \end{block}
  \vspace*{-0.6em}
  \emph{Proof.} The set
  \[ I \defeq \{ i \in \NN \,|\, \neg\mathop{\exists}\limits_{j \in \NN} i < j \times \alpha\,i \leq \alpha\,j
  \} \]
  cannot be in bijection with~$\NN$, as else the~$I$-extracted subsequence
  of~$\alpha$ would not be good. By~\bad{\textsc{lem}}, the set~$I$ is finite.
  Any index~$i_0$ larger than all the indices in~$I$ is a suitable starting
  point for a monotonic subsequence. \qed
  \medskip
  \pause

@@

  \emph{Idea:} Approximate (fictitious) surjections~$\NN \twoheadrightarrow X$
  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. \\ Over time, such an
  approximation can grow to
  \begin{enumerate}
    \item one of $\sigma y$, where~$y \? X$, so it becomes \emph{more defined}; or
    \item one of~$\sigma \tau$, where~$\tau \? X^*$ such that~$a \in
    \sigma\tau$, so it becomes \emph{more surjective}.
  \end{enumerate}
  \medskip
  \pause

@@

  Given a stage-dependent proposition~$P$, $\hil{$\nabla P$}\,\sigma$ expresses that no
  matter how~$\sigma$ evolves to a better approximation~$\tau$,
  eventually~$P\,\tau$ will hold:
  \AgdaHide{
  \begin{code}
    {-# OPTIONS --no-prop --no-import-sorts --allow-unsolved-metas #-}
    open import Agda.Primitive using (Set)
    Prop = Set
    postulate
      X : Set
      X* : Set
      [] : X*
      _∷_ : X* → X → X*
      _⧺_ : X* → X* → X*
      _∈_ : X → X* → Prop
    module Foo where
  \end{code}
  }
  {\small\begin{code}
      data ∇ (P : X* → Prop) (σ : X*) : Prop where
        now : P σ → ∇ P σ
        later₁ : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
        later₂ : (a : X) → ((τ : X*) → a ∈ (σ ⧺ τ) → ∇ P (σ ⧺ τ)) → ∇ P σ
  \end{code}}

  \vspace*{-0.8em}
  For instance, for~$P\,x_0 \ldots x_{n-1} \defeq \exists_{k : \NN} (k < n
  \times x_k = a)$, we have~$\nabla P \, \varepsilon$. \quad ``$\genalpha$ is surjective''
\end{frame}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
\begin{frame}{Slide about toposes}
  (something about overloading of the logical connectives)
\end{frame}}

@@

  \only<7-9>{
    \emph{Does there exist a maximal ideal?}
    \pause
    \good{Yes}, if~$A$ is countable.
    In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable topos}, and
    \emph{bounded first-order consequences} of its existence there \good{do hold} in
    the base.
    \pause
  }

@@

  \jnote{8}{
    J. Roitman, The uses of set theory, \emph{Math.\@ Intelligencer}
    \textbf{14}(1) (1992), 63--69.

@@

  \small
  \begin{multicols}{2}
    For every inhabited set~$X$, \emph{proximally} \\
    there is an enumeration~$\NN \twoheadrightarrow X$.
    \medskip

    A quasiorder is well iff \emph{everywhere}, \\
    every sequence is good.
    \medskip

    A ring element is nilpotent iff \\
    all prime ideals \emph{everywhere} contain it.
    \medskip

    For every ring, \emph{proximally} \\ there is a maximal ideal.
    \medskip

    A relation is well-founded iff \emph{everywhere}, \\ there is no descending chain.
    \medskip

    \emph{Somewhere}, \\ the law of excluded middle holds.
  \end{multicols}
  \pause

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$(X,{\leq})$ be a well quasiorder.
    Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
    is well-founded.
  \end{block}
  \vspace*{-0.5em}

@@

  Unrolling this proof gives a program~$\nabla\mathsf{Good}\,\varepsilon \to
  \prod_{x:X} \mathsf{Acc}\,x$.
\end{frame}

\begin{frame}{The modal type-theoretic multiverse?}
  \emph{As I understand it, \ldots}

  For extensional type theory, similar picture as with toposes:
  Given a model of extensional type theory and a Grothendieck site~$\mathcal{C}$ in it, there
  is a model of extensional type theory deserving the name~``sheaf model
  over~$\mathcal{C}$''. This sheaf model is just a thin translation layer of
  the base model.
  \pause

  For HoTT-like theories, given a ``presheaf model'' we know how to construct a
  ``sheaf model'', but we do not know how to construct presheaf models in the
  first place.
  \pause

>>>>

\newcommand{\dnote}[1]{%
  \begin{tabular}{@{}m{2em}@{}m{0.83\textwidth}@{}}%
    \textdbend &#1%
  \end{tabular}%
  \par
}

@@

  \textbf{Examples.} (classically)\ \ $(\NN,{\leq}),\ \ X \times Y,\ \ X^*,\ \ \mathrm{Tree}(X).$
  \bigskip
  \pause

@@

\begin{frame}{Sequences provided by \textsc{lem}}
  \justifying
  \begin{block}{}
    \justifying
    \textbf{Lemma.} Let~$X$ be a well quasiorder. Let~$\alpha : \NN \to X$ be a
    sequence. Assuming~\bad{\textsc{lem}}, there merely is a monotonic
    subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
  \end{block}
  \vspace*{-0.6em}
  \emph{Proof.} The type
  \[ I \defeq \sum_{i \? \NN} \neg\sum_{j \? \NN} i < j \times \alpha\,i \leq \alpha\,j \]
  cannot be in bijection with~$\NN$, as else the~$I$-extracted subsequence
  of~$\alpha$ would not be good. By~\bad{\textsc{lem}}, the type~$I$ is finite.
  Any index~$i_0$ larger than all the indices in~$I$ is a suitable starting
  point for a monotonic subsequence. \qed
  \medskip
  \pause

@@

  \emph{Idea:} Approximate (fictitious) surjections~$\NN \twoheadrightarrow X$
  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. \\
  Starting with the empty sequence~$\varepsilon$, over time, such an
  approximation can grow to
  \begin{enumerate}
    \item one of $\sigma y$, where~$y \? X$, so it becomes \emph{more defined}; or
    \item one of~$\sigma \tau$, where~$\tau \? X^*$ such that~$a \in
    \sigma\tau$, so it becomes \emph{more surjective}.
  \end{enumerate}
  \medskip
  \pause

@@

  Given a stage-dependent proposition~$P$, $\hil{$\nabla P$}\,\sigma$ expresses that no
  matter how~$\sigma$ evolves to a better approximation~$\tau$,
  eventually~$P\,\tau$ will hold:
  \AgdaHide{
  \begin{code}
    {-# OPTIONS --no-prop --no-import-sorts --allow-unsolved-metas #-}
    open import Agda.Primitive using (Set)
    Prop = Set
    infixl 5 _∈_
    postulate
      X : Set
      X* : Set
      [] : X*
      _∷_ : X* → X → X*
      _⧺_ : X* → X* → X*
      _∈_ : X → X* → Prop
    module Foo where
  \end{code}
  }
  {\small\begin{code}
      data ∇ (P : X* → Prop) (σ : X*) : Prop where
        now : P σ → ∇ P σ
        later₁ : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
        later₂ : (a : X) → ((τ : X*) → a ∈ σ ⧺ τ → ∇ P (σ ⧺ τ)) → ∇ P σ
  \end{code}}

  \vspace*{-0.8em}
  For instance, for~$P_a\,\sigma \defeq (a \in \sigma)$,
  we have~$\nabla P_a \, \varepsilon$. \quad ``$\genalpha$ is surjective''
\end{frame}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{Topos for reifying generic models}
  Grothendieck toposes (= categories of sheaves over sites) are mathematical universes:
  \begin{itemize}
    \item The \hil{generic sequence} is an honest function~$\NN \to X$
    \hil{in} a certain topos~$\mathcal{E}$.
    \item The \hil{generic surjection} is an honest surjection \hil{in} an
    appropriate subtopos of~$\mathcal{E}$.
    \item In the \emph{double-negation subtopos} of the base, a predicate~$Q : \NN
    \to \mathsf{Prop}$ such that
    \[ Q\,x \times Q\,y \longrightarrow x = y \quad\text{and}\quad
      \neg\neg \sum_{n\?\NN} Q\,n \]
    looks like an ordinary natural number.
  \end{itemize}

  A statement holds in a topos iff a certain topos-directed translation holds
  in the base. For instance, for the double-negation subtopos the translation
  substitutes
  \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
    \exists & \leadsto & \exists^\mathrm{cl},
    & \exists^\mathrm{cl} &\defeqv& \neg\neg\exists, \\
    \vee & \leadsto & \vee^\mathrm{cl},
    & \alpha \vee^\mathrm{cl} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
    = & \leadsto & =^\mathrm{cl},
    & s =^\mathrm{cl} t &\defeqv& \neg\neg(s = t).
  \end{array} \]
  In general, stage-dependent~$\nabla$ instead of~$\neg\neg$.
\end{frame}}

@@

  \only<7-9>{
    \emph{Does there exist a maximal ideal?}
    \pause
    \good{Yes}, if~$A$ is countable.
    In the general case: \bad{No}\pause, but \good{yes} in a \emph{suitable topos}, and
    \emph{bounded first-order consequences} of its existence there \good{pass down} to
    the base.
    \pause
  }

@@

  \jnote{8}{
    Judith Roitman. ``The uses of set theory''. \emph{Math.\@ Intelligencer}
    \textbf{14}(1) (1992), 63--69.

@@

  \emph{Traveling the multiverse:}
  \begin{itemize}
    \item For every inhabited set~$X$, \emph{proximally} there is an enumeration~$\NN \twoheadrightarrow X$.
    \item A quasiorder is well iff \emph{everywhere}, every sequence is good.
    \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
    \item For every ring, \emph{proximally} there is a maximal ideal.
    \item A relation is well-founded iff \emph{everywhere}, there is no descending chain.
    \item \emph{Somewhere}, the law of excluded middle holds.
  \end{itemize}
  \pause

  \only<1-3>{
    \small
    \begin{block}{}
      \justifying
      \textbf{Prop.} Let~$(X,{\leq})$ be a well quasiorder.
      Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
      is well-founded.
    \end{block}
    \vspace*{-0.5em}

@@

    Unrolling this proof gives a program~$\nabla\mathsf{Good}\,\varepsilon \to
    \prod_{x:X} \mathsf{Acc}\,x$.
  }
  \pause

  \scriptsize
  \emph{Foreshadowed by:}
  \begin{itemize}
    \item[1984\phantom{s}] André Joyal, Miles Tierney. ``An extension of the Galois theory of Grothendieck''.
    \item[1987\phantom{s}] Andreas Blass. ``Well-ordering and induction in intuitionistic logic and topoi''.
    \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic universes.
    \item[2011\phantom{s}] Joel David Hamkins. ``The set-theoretic multiverse''.
    \item[2013\phantom{s}] Shawn Henry. ``Classifying topoi and preservation of higher order logic by geometric morphisms''.
  \end{itemize}
\end{frame}

\begin{frame}{Towards the modal type-theoretic multiverse}
  \small
  \good{\cmark{}} \emph{There are type-theoretic multiverses,} such as
  \begin{itemize}
    \item the collection of all $\mathrm{PSh}(\mathcal{C} \times \mathcal{B})$,
    where~$\mathcal{B}$ ranges over cube categories and~$\mathcal{C}$ over
    arbitrary small categories, and their corresponding sheaf models

    {\scriptsize
    Coquand. ``A survey of constructive presheaf models of univalence''. \emph{ACM SIGLOG News}, 5.3 (2018).
    \par}
  \end{itemize}
  \medskip
  \pause

  \bad{\xmark} \emph{Accessing the multiverse from within type theory is tricky:}
  \begin{itemize}
    \item Given a model of $\mathfrak{s}$CIC and a category~$\mathcal{C}$ in
    it, we have a syntactic presheaf model of CIC.

    {\scriptsize
    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010).
    Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020). \\
    \par}
    \item Given a suitable lex modality, we have a syntactic sheaf model (model of modal types).

    {\scriptsize
    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021).
    Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016). \\
    \par}
    \item (I believe) we have syntactic sheaf models in certain special cases,
    when no coherence issues arise in defining the notion of presheaves.
  \end{itemize}
